From Real-time Attention Assessment to ``With-me-ness'' in Human-Robot Interaction

### 2015-12-10 paper.tex
Implemented the remaining reviewers' comments + fit into 8 pages

Measuring ``how much the human is in the interaction'' -- the \emph{level of
engagement} -- is instrumental in building effective interactive robots.

Engagement, however, is a complex, multi-faceted cognitive mechanism that is
only indirectly observable. This article formalizes \emph{with-me-ness} as one
of such indirect measures. \emph{With-me-ness}, a concept borrowed from the
field of {\it Computer-Supported Collaborative Learning}, measures in a
well-defined way to what extent the human is \emph{with} the robot over the
course of an interactive task. As such, it is a meaningful precursor of
engagement.

We expose the full methodology, from real-time estimation of the human's focus
of attention (relying on a novel, open-source, vision-based head pose
estimator), to on-line computation of with-me-ness.

We report as well on the experimental validation of this approach, using a
naturalistic setup involving children during a complex robot-teaching task.

@@

@@

As explained, we derive the visual field of attention from the head pose. Our
technique only involves a single monocular RGB camera used for facial feature
extraction, and a static simplified 3D mesh of a human head.  68 facial features
are extracted using a fast template-based face alignment algorithm by Kazemi and
Sullivan~\cite{kazemi2014one}, as implemented in the open-source {\tt dlib}
library~\cite{dlib09}.  Eight of these features are then matched to their 3D
counterparts (Figure~\ref{head_pose}) and we rely on an iterative $PnP$
algorithm (OpenCV's implementation) to compute the translation and rotation of
the head with respect to the camera frame. With this approach, knowing the
intrinsic parameters of the camera (calibrated camera) is required for an
accurate estimation of the absolute 3D localization of the head.

@@

We model the field of attention as the central region of the field of view.  The
field of view itself is approximated to a cone spanned from the nasal depression
(sellion) of the human face. Different dimensions for the human field of view
can be found in the litterature: Holmqvist~\cite{holmqvist2011eye} models it
with an horizontal aperture of $ \pm 40\degree $ and a vertical aperture of $
\pm 25\degree $, while Spector~\cite{walker1980clinical} for instance suggests
$60\degree$ up, $75\degree$ down, $60\degree$ inwards (towards the nose) and
$95\degree$ outwards.  Previous work on visual perspective taking for social
robotics~\cite{sisbot2011situation} model the field of attention as a cone of
$30\degree$. We retained in this work a slightly wider aperture of $40\degree$.

We then approximate the visual \emph{focus} of attention (VFoA) of the human to
the objects which lie inside this field of attention (Figure~\ref{fig:vfoa}). At
a given time, more than one object can therefore be \emph{in focus}.

@@

Within these limitations, computing if object $A(x_A,y_A,z_A)$ is in the field
of attention of the human requires first to transform the coordinates
$A(X_A,Y_A,Z_A)$ into the frame of the face, and then to verify the following
simple inequality (with $fov$ the aperture, and assuming that the main axis
$\vec{x}$ of the field of attention points forward):

\begin{align}
    \sqrt{Y_A^2 + Z_A^2} < tan\left(\frac{fov}{2}\right) \cdot X_A
\label{eq:fov}
\end{align}

@@

The experiment, part of the CoWriter project~\cite{Hood:2015}, involves a robot
which tries to engage a child in handwriting tasks using a \emph{learning by
teaching} paradigm (\ie the child is the teacher, and he/she attempts to improve
the robot's handwriting). A tactile tablet is used as writing support.

Figures~\ref{fig:setup} and~\ref{fig:realSetup} illustrate the experimental
setup: a face-to-face child-robot interaction with an (autonomous) Aldebaran
{\sc nao} robot, in the presence of a facilitator (one of the researchers).

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\columnwidth]{experimental_setup}
    \caption{\small \textbf{Experimental setup}: face-to-face interaction with a {\sc
            nao} robot. The robot writes on the tactile tablet, the child then
            corrects the robot by directly overwriting the words on the tablet
            with a stylus. The facilitator remains next to the child to guide the work. 
            The secondary tablet allows the child to tell the robot what to
            write. The areas of interest -- corresponding to potential target of
            attention -- are circled in red with their name.}
    \label{fig:setup}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\columnwidth]{realSetup}
    \caption{\small Picture of the interaction with one of the children.}
    \label{fig:realSetup}
\end{figure}

@@

The interaction is organized in rounds of writing: during a typical round, the
child requests the robot to write something (a single letter, a number, or a
full word), and presents a tactile tablet (equipped with a custom writing
application) to the robot. The robot writes on the tablet by mimicking the
writing gesture (but without actually physically touching the tablet); the child
then pulls back the tablet, corrects the robot's attempt by writing on top of or
next to the robot's writing, and ``sends'' his/her demonstration to the robot by
pressing a small button on the tablet. The robot learns from this demonstration
and tries again. The child continues the turn-taking until they decide to train
the robot on another word. In total, the children performed on average 12.16
($SD=2.61$) rounds of writing (complete details on the rationale and
implementation of this experiment can be found in~\cite{Hood:2015}).

Once per interaction, the robot interrupts the handwriting task to tell a story
(taking about 2 min), and the turn-based hand-writing task continues afterwards.
The intended purpose of the story-telling episode is to locally trigger a
different set of attention behaviors for the child.

@@

    \node at (0,0) [cmpt, fill=Green!50] (tf) {tf};
    \node [cmpt, fill=YellowOrange!50, left=of tf] (attention) {attention tracking};
    \node [above=0.6cm of attention, text width=2.7cm,align=center] (input) {\tt /image \\ /camera\_info \\ dlib {\sf faces model} \\ {\sf 3D face template}};
    \node [cmpt, right=of tf] (static) {static pose publishers};
    \node [above=0.6cm of static, text width=2.7cm,align=center] (staticposes) {\sf locations of \\areas of interest};
    \node [cmpt, above=0.6cm of tf] (pose) {robot state publisher};
    \node [cmpt, fill=YellowOrange!50, below=0.6cm of tf] (focus) {focus estimation};
    \node [below=0.6cm of focus] (output) {\tt /frames\_in\_focus};

@@

\end{tikzpicture}
}

    \caption{\small ROS nodes involved in the VFoA estimation (orange nodes were developed for this work).}
    \label{fig:system}
\end{figure}

The experiment was carried out with an Aldebaran {\sc nao} robot, using ROS as a
middleware to build the attention estimation pipeline (Figure~\ref{fig:system}).
Head pose estimation, presented in section~\ref{sec:vfoa}, builds on the {\tt
dlib} and OpenCV libraries; the pose transformations are handled by the ROS {\sc
tf} library. The same {\sc tf} library is used to represent the possible point
of interests as individual frames: an object is considered to be in focus when
its frame lies within the field of attention of the participant
(Figure~\ref{fig:vfoa}).  The implementation is open-source and available at
\url{https://github.com/chili-epfl/attention-tracker}.

@@

\begin{table}[h!]
    \centering
    \caption{\textbf{Levels of with-me-ness}. For each subject, the with-me-ness
    level is reported over the whole interaction, either based on the
    ground-truth (\ie \emph{ground-truth with-me-ness}), or based on the focus of
    attention measured by the robot.}

@@

Looking at the ground-truth focus of attention (orange lines), the first
striking observation is that this child did generally \emph{closely} follow what
the robot was expecting: in that regard, it seems that the child was very much
engaged in the interaction (we discuss in the next section the exact
relationships between with-me-ness and engagement). The only major exception is
the story-telling phase (episode ``D''): the child was seemingly not captivated
by the first half of the story, and their attention was not directed towards the
robot.

@@

Interestingly, previous research conducted on ape-human interaction by
Tomasello~\etal~\cite{tomasello2007reliance} shows that apes rely primarily on
the head pose to estimate the focus of attention of a human partner, in to
contrast to human children who primarily rely on eye direction. From this
perspective, our robot mimics attention tracking mechanism comparable to the
ones found in apes.

@@

Compared to engagement, one of the strengths of with-me-ness is its specificity:
it is well-defined (the only important variable -- that needs to be reported --
is the mapping {\it phase/expected attentional target}, \ie our
Table~\ref{tab:targets}), we can formalize it, and as such, it is valuable to
assess and compare how users are willing or able to interact with a robot.  We
have hopefully demonstrated in this article that with-me-ness is an operational
\emph{in-the-moment} metric that can be also used as feedback to the robot
controller to build richer, more adaptive interactive behaviors for our robots.

>>>>

Measuring ``how much the human is in the interaction'' -- the \emph{level of
engagement} -- is instrumental in building effective interactive robots.
Engagement, however, is a complex, multi-faceted cognitive mechanism that is
only indirectly observable. This article formalizes \emph{with-me-ness} as one
of such indirect measures. \emph{With-me-ness}, a concept borrowed from the
field of {\it Computer-Supported Collaborative Learning}, measures in a
well-defined way to what extent the human is \emph{with} the robot over the
course of an interactive task. As such, it is a meaningful precursor of
engagement.  We expose in this paper the full methodology, from real-time
estimation of the human's focus of attention (relying on a novel, open-source,
vision-based head pose estimator), to on-line computation of with-me-ness.  We
report as well on the experimental validation of this approach, using a
naturalistic setup involving children during a complex robot-teaching task.

@@

\begin{IEEEkeywords}
Human-Robot Interaction; Visual Focus of Attention; With-me-ness; Real-time Head Pose Estimation.
\end{IEEEkeywords}

@@

As explained, we derive the visual field of attention from the head pose. Our
technique only involves a single monocular RGB camera used for facial feature
extraction, and a static simplified 3D mesh of a human head.  68 facial features
are extracted using a fast template-based face alignment algorithm by Kazemi and
Sullivan~\cite{kazemi2014one}, as implemented in the open-source {\tt dlib}
library~\cite{dlib09}.  Eight of these features (chosen to be far apart and
relatively stable across age and gender) are then matched to their 3D
counterparts (Figure~\ref{head_pose}) and we rely on an iterative $PnP$
algorithm (OpenCV's implementation) to compute the translation and rotation of
the head with respect to the camera frame. With this approach, knowing the
intrinsic parameters of the camera (calibrated camera) is required for an
accurate estimation of the absolute 3D localization of the head.

@@

We model the field of attention as the central region of the field of view.  The
field of view itself is approximated to a cone spanned from the nasal depression
(sellion) of the human face. Different dimensions for the human field of view
can be found in the literature: Holmqvist~\cite{holmqvist2011eye} models it
with an horizontal aperture of $ \pm 40\degree $ and a vertical aperture of $
\pm 25\degree $, while Walker~\cite{walker1980clinical} for instance suggests
$60\degree$ up, $75\degree$ down, $60\degree$ inwards (towards the nose) and
$95\degree$ outwards.  Previous work on visual perspective taking for social
robotics~\cite{sisbot2011situation} model the field of attention as a cone of
$30\degree$. We retained in this work a slightly wider aperture of $40\degree$.
We then approximate the visual \emph{focus} of attention (VFoA) of the human to
the objects which lie inside this field of attention (Figure~\ref{fig:vfoa}). At
a given time, more than one object can therefore be \emph{in focus}.

@@

Within these limitations, computing if object $A(x_A,y_A,z_A)$ is in the field
of attention of the human requires first to transform the coordinates
$A(X_A,Y_A,Z_A)$ into the frame of the face, and then to verify the
simple inequality $\sqrt{Y_A^2 + Z_A^2} < tan\left(\frac{fov}{2}\right) \cdot
X_A$ (with $fov$ the aperture, and assuming that the main axis
$\vec{x}$ of the field of attention points forward).

@@

The experiment, part of the CoWriter project~\cite{Hood:2015}, involves a robot
which tries to engage a child in handwriting tasks using a \emph{learning by
teaching} paradigm (\ie the child is the teacher, and he/she attempts to improve
the robot's handwriting). A tactile tablet is used as writing support.
Figures~\ref{fig:setup} and~\ref{fig:realSetup} illustrate the experimental
setup: a face-to-face child-robot interaction with an (autonomous) Aldebaran
{\sc nao} robot, in the presence of a facilitator (one of the researchers).

\begin{figure}[ht!]
    \centering
    \includegraphics[width=0.8\columnwidth]{experimental_setup}
    \caption{\small \textbf{Experimental setup}: face-to-face interaction with a {\sc
            nao} robot. The robot writes on the tactile tablet, the child then
            corrects the robot by directly overwriting the words on the tablet
            with a stylus. The facilitator remains next to the child to guide the work. 
            The secondary tablet allows the child to tell the robot what to
            write. The areas of interest -- corresponding to potential target of
            attention -- are circled in red with their name.}
    \label{fig:setup}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics[width=1\columnwidth]{realSetup}
    \caption{\small Picture of the interaction with one of the children.}
    \label{fig:realSetup}
\end{figure}

@@

The interaction is organized in rounds of writing: during a typical round, the
child requests the robot to write something (a single letter, a number, or a
full word), and presents a tactile tablet (equipped with a custom writing
application) to the robot. The robot ``writes'' on the tablet by drawing in the
air the letters that are displayed on the screen by the tablet application; the child
then pulls back the tablet, corrects the robot's attempt by writing on top of or
next to the robot's writing, and ``sends'' his/her demonstration to the robot by
pressing a small button on the tablet. The robot learns from this demonstration
and tries again. The child continues the turn-taking until they decide to train
the robot on another word. In total, the children performed on average 12.16
($SD=2.61$) rounds of writing (complete details on the rationale and
implementation of this experiment can be found in~\cite{Hood:2015}).

Once per interaction, the robot interrupts the handwriting task to tell a story
(taking about 2 min), and the turn-based hand-writing task continues afterwards.
The intended purpose of the story-telling episode is to break the routine of the
writing turns by creating a surprise, and thus, to elicit a different set of
attention behaviors from the child.

@@

    \node at (0,0) [cmpt, fill=Green!50] (tf) {tf};
    \node [cmpt, fill=YellowOrange!50, left=of tf] (attention) {attention tracking};
    \node [above=0.6cm of attention, text width=3cm,align=center] (input) {{\tt
    /image}, {\tt /camera\_info}, \\ {\tt dlib} {\sf faces model}, \\ {\sf 3D face template}};
    \node [cmpt, right=of tf] (static) {static pose publishers};
    \node [above=0.6cm of static, text width=2.7cm,align=center] (staticposes) {\sf locations of \\areas of interest};
    \node [cmpt, above=0.6cm of tf] (pose) {robot state publisher};
    \node [cmpt, fill=YellowOrange!50, below=0.6cm of tf] (focus) {focus estimation};
    \node [below=0.6cm of focus] (output) {\tt /frames\_in\_focus};

@@

\end{tikzpicture}
}
    \caption{\small ROS nodes involved in the VFoA estimation (orange nodes were
    specifically developed for this work).}
    \label{fig:system}
\end{figure}

The experiment was carried out with an Aldebaran {\sc nao} robot, using ROS as a
middleware to build the attention estimation pipeline (Figure~\ref{fig:system}).
Head pose estimation, presented in section~\ref{sec:vfoa}, builds on the {\tt
dlib} and OpenCV libraries; the pose transformations are handled by the ROS {\sc
tf} library. The same {\sc tf} library is used to represent the possible point
of interests as individual frames: an object is considered to be in focus when
its frame lies within the field of attention of the participant
(Figure~\ref{fig:vfoa}).  The implementation is open-source and available at
\texttt{\url{https://github.com/chili-epfl/attention-tracker}}.

@@

\begin{table}[h!]
    \centering
    \caption{\textbf{Levels of with-me-ness}. For each subject, the with-me-ness
    level is reported over the whole interaction, either based on the
    annotated focus of attention (\ie \emph{ground-truth with-me-ness}), or based on the focus of
    attention measured by the robot.}

@@

Looking at the ground-truth focus of attention (orange lines), the first
striking observation is that this child did generally \emph{closely} follow what
the robot was expecting: in that regard, it seems that the child was very much
engaged in the interaction (we discuss in the next section the exact
relationships between with-me-ness and engagement). The only major exception is
the story-telling phase (episode ``D''): the child was seemingly not captivated
by the first half of the story, and their attention was not directed towards the
robot (this actually matches the observed behavior of the children who mostly
found the story boring).

@@

%Interestingly, previous research conducted on ape-human interaction by
%Tomasello~\etal~\cite{tomasello2007reliance} shows that apes rely primarily on
%the head pose to estimate the focus of attention of a human partner, in to
%contrast to human children who primarily rely on eye direction. From this
%perspective, our robot mimics attention tracking mechanism comparable to the
%ones found in apes.

@@

Compared to engagement, one of the strengths of with-me-ness is its specificity:
it is well-defined, we can formalize it, and as such, it is valuable to
assess and compare how users are willing or able to interact with a robot.  We
have hopefully demonstrated in this article that with-me-ness is an operational
\emph{in-the-moment} metric that can also be used as a real-time feedback to the
robot controller to build richer, more adaptive interactive behaviors for our
robots.

Note however that, besides the actual focus of attention, the mapping {\it
phase/expected attentional target} (\ie our Table~\ref{tab:targets}) is a
critical piece of information to interpret with-me-ness. The mapping is
typically built by a domain expert, and is often subject to debate (for
instance in our experiment, one could argue that during the ``Waiting for
feedback'' phase, the child could have gazed toward the robot to make sure the
robot was paying attention, and consequently, {\sf robot} should be added to the
expected target).  For this reason, the chosen mapping should always be reported
along with the computed with-me-ness levels, and with-me-ness should not be
reported as an absolute metric, but rather as a mean of comparing
different interactions within the same study.
